---
title: "DSC 475 - Project 1"
author: "Hao Ma"
date: "Fall 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **1.**
```{r}
library(readxl)
data <- read_excel("Project1_DataSet.xlsx")
plot(data$`Miles, in Millions`, main="Monthly Data on Airline Miles",
     xlab="Time (Jan 64 - Dec 70)", ylab="Miles (Millions)")
lines(data$`Miles, in Millions`)
```


## **2.**
```{r}
acf(data$`Miles, in Millions`, main="ACF Plot", lag.max=48)
```
**The seasonal period is 12 months.**


## **3.**
```{r}
# A function to compute simple moving average with a given window length
# and a univariate dataset. Then plot the data along with the SMA results. 
SMA <- function(N, data){
  T_ <- nrow(data)
  t <- N:T_
  M <- c()
  for(i in t){
    M <- c(M, sum(data[,2][(i-N+1):i])/N)
  }
  plot(data[,2], xlab="Time Index", ylab="Miles (millions)")
  lines(data[,2])
  points(t, M, pch=2, col="red")
  lines(t, M, pch=2, col="red")
  legend("bottomright", c("Actual","Smoothed"), 
         col=c("black","red"), pch=c(1,2))
}
```

```{r}
SMA(10, as.data.frame(data))
```

```{r}
SMA(20, as.data.frame(data))
```

```{r}
SMA(30, as.data.frame(data))
```

```{r}
SMA(50, as.data.frame(data))
```
**I chose 20 as the window length based on the plots shown above. We do not want the smoothed curve be "too smooth" like a straight line or "too accurate" like the original data. Somewhere in the middle is sufficient to discover the trend.**


## **4.**
**It has an increasing trend.**


## **5.**
```{r}
# Then compute first difference to remove trend
diffed <- diff(data$`Miles, in Millions`, differences=1)
acf(diffed, lag.max=48, main="ACF for Differenced Data")
pacf(diffed, lag.max=48, main="PACF for Differenced Data")
```
**There are many significant lags due to the seasonality within each year (non-stationarity).**


## **6.**
```{r}
# Compute first seasonal difference to remove seasonality as well with 12 as 
# the seasonal period
seasonal_diff <- diff(diffed, lag=12, differneces=1)
acf(seasonal_diff, lag.max=48, main="ACF for First Seasonal Difference")
pacf(seasonal_diff, lag.max=48, main="PACF for First Seasonal Difference")
```
**The number of significant lags decreases and they are all within the first year.**


## **7.**
Based on the trend, seasonality and auto-correlation plots, we will develop a SARIMA model using the  `auto.arima()` function in the `forecast` library. Set $d=1$, $D=1$ and vary $p,q,P,Q$ each over the range 0 to 3 to find the best model based on BIC.  
```{r}
library(forecast)

# Use the first 6 years of monthly data to create a time series object
training <- ts(data[1:72,2], start=c(1964, 1), frequency=12)
# Then search for the best combination of parameters using `auto.arima()`
model <- auto.arima(training, d=1, D=1, max.p=3, max.q=3, max.P=3, max.Q=3,
           start.p=0, start.q=0, start.P=0, start.Q=0, ic="bic", trace=T)
```
**The best model is:** $ARIMA(0,1,2)(1,1,0)_{12}$.


## **8.**
Use the model above to forecast for the year 1970 (12 forecasts) using the function `forecast()`.
```{r}
my_forecast <- forecast(model, h=12)
my_forecast
```
The forecasts and prediction intervals are shown above.

```{r}
# Compare the mean forecasts with the actual data
actual <- data$`Miles, in Millions`[73:84]
forecasts <- as.numeric(my_forecast$mean)
plot(actual, main="Forecasts and Actual Values for 1970",
     ylim=c(9,17), xlab="Time (Jan-Dec 1970)", ylab="Miles (millions)")
lines(actual)
points(forecasts, col="red", pch=2)
lines(forecasts, col="red", pch=2)
legend("topleft", c("Actual","Forecast"), 
       col=c("black","red"), pch=c(1,2))
```
The monthly trend and values of the forecast are close to the actual data, except for the time period of April 1970 where the decrease is too steep. This should be acceptable.
